Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-04 paper/main.tex
results

\renewcommand{\subsubsection}[1]{\smallskip{}\noindent\textbf{#1}.}

@@

\subsubsection{UMM}
%
UMM uses the same budget of evaluations ($\FEmax$) as CEGO and initially
samples $\minit=10$ solutions uniformly.
%At each sampling step, we set
%$\beta=10$, meaning that 10 permutations are generated and the farthest from
%$S$ is selected for evaluation.
The other two parameters of UMM are $r_1$ and $r_2$, which determine the
balance between exploration and exploitation. In the experimental section we
will study their impact.

\subsubsection{CEGO} \EKHINE{tenemos una pagina entera dedicada a CEGO :s}\MANUEL{Sugiere qué quitar usando el commando \textbackslash{}ekhine}
Combinatorial Efficient Global Optimization (CEGO)~\citep{ZaeStoFriFisNauBar2014} is an extension of the well-known EGO
method~\citep{JonSchWel98go} to unconstrained black-box combinatorial
optimization problems. In EGO, Gaussian process models are used as a surrogate
of the landscape of the expensive original problem. An optimization method
searches for solutions in the surrogate model by optimizing the expected
improvement criterion, which balances the expected mean and variance of the
chosen solution. Once a solution is chosen, it is evaluated on the actual
fitness function and the result is used to update the surrogate-model,
hopefully increasing its predictive power.

@@

\subsubsection{Other settings}
%
For simplicity, we use Kendall's-$\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all experiments.
%Nevertheless,
%\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
%when using Hamming distance and the same may apply to UMM.
We plan to extend UMM to other distance measures in the future.  In all
experiments, we consider a maximum budget of $\FEmax=400$ evaluations of the
actual objective function. In a white-box context, state-of-the-art algorithms
for the LOP and PFSP typically evaluate thousands of solutions,
thus, the budget considered here for the black-box context is extremely
limited.

\subsubsection{Benchmark problems}
%
Experiments with real-world expensive black-box problems would be computationally infeasible. Instead, we consider classical combinatorial optimization problems as black-box optimization problems. For consistency with the rest of the paper, a permutation $\sigma$ denotes a ranking also in this section, thus we use the inverse permutation $\sigma^{-1}$ in the definition of the problems. Later in the experimental section we discuss the effects of not performing this inversion.
%
The \emph{Linear Ordering Problem (LOP)}, when minimizing, is defined as
%\footnote{For consistency with the other problems, we minimise the lower triangle instead of maximising the upper triangle.} 
\begin{equation}\label{eq:lop}
  \min_{\sigma \in S_n} f(\sigma) = \sum_{i=1}^{n} \sum_{j=1}^{i-1}  a_{\sigma^{-1}(i), \sigma^{-1}(j)}
\end{equation}
%
where $[a_{i,j}]$ is a matrix of size $n \times n$.

@@

\subsubsection{Datasets}
%
%For the LOP, we generated synthetic instances as follows. First, we create a sample $S$ of 200 permutations of size $n=20$ from a MM with $\sigma_0$ equal to the identity permutation and $\theta=-\log(\textit{Uniformity})$, where  $\textit{Uniformity}\in\{0.5,$ $0.7,$ $0.9\}$, i.e., from peaked distributions to almost uniform distributions. Then, we generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbb{I}\{\sigma(i)<\sigma(j)\}$, where ${\mathbb{I}\colon S \times S \to \{0,1\}}$ is the indicator function. In principle, these should be best-case instances for UMM since their fitness landscape closely matches the Mallows model.%\MANUEL{Ekhine: Could you revise this text? I wanted to explain why we use synthetic datasets.}
We consider real LOP datasets from the LOLIB
repository.\footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}} In
particular, we have taken two instances of various types focusing on relatively
small instances, since he computation time of CEGO grows rapidly with instance
size: \texttt{N-p40-01}, \texttt{N-p40-02}, \texttt{N-p50-01} and
\texttt{N-p50-02} of type \emph{RandB}; \texttt{N-t59d11xx} and
\texttt{N-t59b11xx} of type \emph{IO}; and \texttt{N-sgb75.01} and
\texttt{N-sgb75.02} of type \emph{SGB}.
% Random instances of type B of various sizes $n \in \{50, 100, 150, 200,
% 250\}$. These are harder problems than the synthetic instances, not only
% because of the larger search space, but also because of the different instance
% types.
In the case of the PFSP, we consider the same instances as
\citet{ZaeStoBar2014:ppsn}, i.e., \texttt{reC05}, \texttt{reC13},
\texttt{reC19} and \texttt{rec31}, with $n \in \{20, 20, 30, 50\}$ and $M \in \{5, 20, 10, 15\}$, respectively.

@@

\subsubsection{Computing environment.}
% Due to the computational effort required, we used several computing systems to
% run experiments.
% All UMM experiments were either on run on Intel Xeon
% \emph{Broadwell} E5-2683v4 CPUs at 2.10\,GHz or Intel Xeon \emph{Haswell} E5-2680v3
% CPUs at 2.50\,GHz, both of them with 32\,GB RAM. 
 %
Experiments were run on Intel Xeon \emph{Ivybridge} E5-2650v2 CPUs at
2.60\,GHz, 64\,GB RAM running CentOS Linux release 7 (Core).
%The operating system was CentOS Linux release 7 (Core).

@@

For each problem instance and each algorithm, we plot the mean fitness of the
solution evaluated at each step over 10 runs with different random
seeds, as shown in Fig.~\ref{fig:results}.
%
The shaded area shows one standard deviation around the mean. Ideally, each new
evaluation will monotonically improve in fitness, since each solution evaluated
helps to refine the model and leads to a better solution being evaluated in the
next step.  However, it is possible that new solutions are worse than their
predecessors. %Nevertheless, in this example, there is a clear
% overall improvement as more evaluations are performed.

Figure~\ref{fig:results} shows the results on the real-world instances from LOLIB and PFSP. There is one figure per instance. In each figure there is one horizontal line displaying the best known solutions for the problem. Besides, there are 5 different lines representing each of the different algorithm/versions considered, in particular:

@@

\begin{figure*}[tb]
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-sgb75-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p50-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p50-01}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\end{figure*}

@@

\subsubsection{Reproducibility} Source code, datasets and scripts necessary to reproduce the results are available at \supplement.
%  \vspace*{-1em}
%\begin{acks}
%% MANUEL: Blind, so no acknowledgments yet.
%\paragraph*{Acknowledgements.} Blind
% M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
%\manuel{}{: We should also thank the COST ACTION: Ask Carlos?} \ekhine{}{Sure!, esto no habra q ponerlo aun por ser blind, no?}
%Thanks to: Hao Wang (Leiden University) for pointing us to the arguments of
%\citep{EriPeaGar2019scalable}; Thomas Stützle (Université libre de Bruxelles) for suggesting works
%about differences between QAPLIB instances; and Ahmed Boujaada for the experimental support. 
  % supported by the \grantsponsor{GS501100001809}{National Natural
  %   Science Foundation of
  %   China}{http://dx.doi.org/10.13039/501100001809} under Grant
  % No.:~\grantnum{GS501100001809}{61273304}
  % and~\grantnum[http://www.nnsf.cn/youngscientists]{GS501100001809}{Young
  %   Scientists' Support Program}.
%\end{acks}

>>>>

\renewcommand{\myparagraph}[1]{\smallskip{}\noindent\textbf{#1}.}

@@

\myparagraph{UMM}
%
UMM uses the same budget of evaluations ($\FEmax$) as CEGO and initially
samples $\minit=10$ solutions uniformly.
%At each sampling step, we set
%$\beta=10$, meaning that 10 permutations are generated and the farthest from
%$S$ is selected for evaluation.
The other two parameters of UMM are $r_1$ and $r_2$, which determine the
balance between exploration and exploitation. In the experimental section we
will study their impact.

\myparagraph{CEGO} \EKHINE{tenemos una pagina entera dedicada a CEGO :s}\MANUEL{Sugiere qué quitar usando el commando \textbackslash{}ekhine}
Combinatorial Efficient Global Optimization (CEGO)~\citep{ZaeStoFriFisNauBar2014} is an extension of the well-known EGO
method~\citep{JonSchWel98go} to unconstrained black-box combinatorial
optimization problems. In EGO, Gaussian process models are used as a surrogate
of the landscape of the expensive original problem. An optimization method
searches for solutions in the surrogate model by optimizing the expected
improvement criterion, which balances the expected mean and variance of the
chosen solution. Once a solution is chosen, it is evaluated on the actual
fitness function and the result is used to update the surrogate-model,
hopefully increasing its predictive power.

@@

\myparagraph{Other settings}
%
For simplicity, we use Kendall's-$\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all experiments.
%Nevertheless,
%\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
%when using Hamming distance and the same may apply to UMM.
We plan to extend UMM to other distance measures in the future.  In all
experiments, we consider a maximum budget of $\FEmax=400$ evaluations of the
actual objective function. In a white-box context, state-of-the-art algorithms
for the LOP and PFSP typically evaluate thousands of solutions,
thus, the budget considered here for the black-box context is extremely
limited.

\myparagraph{Benchmark problems}
%
Experiments with real-world expensive black-box problems would be computationally infeasible. Instead, we consider classical combinatorial optimization problems as black-box optimization problems. For consistency with the rest of the paper, a permutation $\sigma$ denotes a ranking also in this section, thus we use the inverse permutation $\sigma^{-1}$ in the definition of the problems. Later in the experimental section we discuss the effects of not performing this inversion.
%
The \emph{Linear Ordering Problem (LOP)}, when minimizing, is defined as
%\footnote{For consistency with the other problems, we minimise the lower triangle instead of maximising the upper triangle.} 
\begin{equation}\label{eq:lop}
  \min_{\sigma \in S_n} f(\sigma) = \sum_{i=1}^{n} \sum_{j=1}^{i-1}  a_{\sigma^{-1}(i), \sigma^{-1}(j)}
\end{equation}
%
where $[a_{i,j}]$ is a matrix of size $n \times n$.

@@

\myparagraph{Datasets}
%
%For the LOP, we generated synthetic instances as follows. First, we create a sample $S$ of 200 permutations of size $n=20$ from a MM with $\sigma_0$ equal to the identity permutation and $\theta=-\log(\textit{Uniformity})$, where  $\textit{Uniformity}\in\{0.5,$ $0.7,$ $0.9\}$, i.e., from peaked distributions to almost uniform distributions. Then, we generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbb{I}\{\sigma(i)<\sigma(j)\}$, where ${\mathbb{I}\colon S \times S \to \{0,1\}}$ is the indicator function. In principle, these should be best-case instances for UMM since their fitness landscape closely matches the Mallows model.%\MANUEL{Ekhine: Could you revise this text? I wanted to explain why we use synthetic datasets.}
We consider real LOP datasets from the LOLIB
repository.\footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}} In
particular, we have taken two instances of various types focusing on relatively
small instances, since he computation time of CEGO grows rapidly with instance
size: \texttt{N-p40-01}, \texttt{N-p40-02}, \texttt{N-p50-01} and
\texttt{N-p50-02} of type \emph{RandB}; \texttt{N-t59d11xx} and
\texttt{N-t59b11xx} of type \emph{IO}; and \texttt{N-sgb75.01} and
\texttt{N-sgb75.02} of type \emph{SGB}.
% Random instances of type B of various sizes $n \in \{50, 100, 150, 200,
% 250\}$. These are harder problems than the synthetic instances, not only
% because of the larger search space, but also because of the different instance
% types.
In the case of the PFSP, we consider the same instances as
\citet{ZaeStoBar2014:ppsn}, i.e., \texttt{reC05}, \texttt{reC13},
\texttt{reC19} and \texttt{rec31}, with $n \in \{20, 20, 30, 50\}$ and $M \in \{5, 20, 10, 15\}$, respectively.

@@

\myparagraph{Computing environment.}
% Due to the computational effort required, we used several computing systems to
% run experiments.
% All UMM experiments were either on run on Intel Xeon
% \emph{Broadwell} E5-2683v4 CPUs at 2.10\,GHz or Intel Xeon \emph{Haswell} E5-2680v3
% CPUs at 2.50\,GHz, both of them with 32\,GB RAM. 
 %
Experiments were run on Intel Xeon \emph{Ivybridge} E5-2650v2 CPUs at
2.60\,GHz, 64\,GB RAM running CentOS Linux release 7 (Core).
%The operating system was CentOS Linux release 7 (Core).

@@

Figure~\ref{fig:results} shows such plots for a few LOP instances from LOLIB
and the PFSP.  For each problem instance and each algorithm, we plot the mean
fitness of the solution evaluated at each step over 10 runs with different
random seeds. The horizontal line indicates the best-known fitness for that instance.
% 
The shaded area shows one standard deviation around the mean. Ideally, each new
evaluation will monotonically improve in fitness, since each solution evaluated
helps to refine the model and leads to a better solution being evaluated in the
next step.  However, it is possible that new solutions are worse than their
predecessors. %Nevertheless, in this example, there is a clear
% overall improvement as more evaluations are performed.

\myparagraph{Convergence analysis of UMM} We start our analysis with the three
configurations of UMM. We can see a common trend in all instances: After
the 10 random initial permutations, the fitness of new solutions decreases
with a linear trend along iterations. The (close to) linear decrease is given
by the linear decrease in the expected distance\MANUEL{distance to the sampling model or variance of the sampling model?} of the sampling model, as described in Section~\ref{sec:sampling}. Overall, the rate of improvement
does not seem to converge, not even for the longer runs of 400 evaluations. Therefore, we conjecture that longer runs are likely to produce further improved solutions. Moreover, a non-linear decrease in the expected distance should result in a faster convergence for larger evaluation budgets. \manuel{The
average runtime for all the instances is around 0.3 minutes, 1.15 minutes and
4.5 minutes respectively for the versions with 100, 200 and 400 fitness
evaluations.}{}\MANUEL{We should discuss the time when discussing the tables.}

\paragraph{Convergence analysis of CEGO}
The red and the gold lines in the plots shown in Fig.~\ref{fig:results}
correspond to the two variants of CEGO the original \CEGOorig and our modified
version \CEGOinv. The different between these two variants is overwhelmingly in
favour of \CEGOinv. With the exception of the \texttt{reC19} PFSP instance, the
search behavior of \CEGOorig resembles a random search, after a brief
improvement phase at the start of the run. By contrast, \CEGOinv shows a very
fast improvement in fitness within the first 50 evaluations that tends to
converge around the 150$^\text{th}$ evaluation, showing little improvement
after that. Since the only difference between these two variants is the
transformation from rankings to orderings before evaluating the objective
function (Sec.~\ref{sec:exper_ro}), we must conclude that this difference has
tremendous effect in performance. Therefore, we will focus our analysis on
\CEGOinv.

\myparagraph{Comparison of UMM and CEGO}

When comparing the UMM and CEGO configurations shown in Fig.~\ref{fig:results},
we can observe that all UMM configurations perform much better than the
original CEGO (\CEGOorig), while they converge slower than our modified
\CEGOinv. However, the rate of improvement of the UMM variants does not seem to
slow down near the maximum budget given and, in several instances such as
\texttt{rec31} and \texttt{N-sgb75-01}, matches and surpasses the solutions
found by \CEGOinv. Interestingly, this is the case, even for the shorter budget
configurations of UMM, i.e., UMM$_{100}$ and UMM$_{200}$. We also observe a
pattern that UMM performs better, or \CEGOinv performs worse, on larger
instances, as can be seen by comparing \texttt{rec19} with \texttt{rec31}.

\myparagraph{Comparison of the best solutions found} We now turn our analysis
to the best (minimum) fitness found at the end of the run by each algorithm. We
show in Table~\ref{tab:results}

For the reason for the abrupt change between both version for
every instance, we refer the reader to Section~\ref{sec:exper_ro}. For the
original version, there is a global decrease in the fitness function. However,
a closer analysis of each iteration does not show a real convergence to any
point. However, with the modified version in which rankings are modeled and
orderings are evaluated we can appreciate the a decrease in terms of fitness
per evaluation. Overall, the modified CEGO has a very abrupt decrease at the
first 50 iterations. At one point, the algorithm seems to have converged and
very little improvement is don from the 150-th iteration onwards.  The average
running time is approximately 22 hours for the 400 function evaluations, which
is 300 times slower than UMM for the same number of function evaluations.

@@

\begin{figure*}[tb]
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p50-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p50-01}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-sgb75-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\end{figure*}

@@

\myparagraph{Reproducibility} Source code, datasets and scripts necessary to reproduce the results are available at \supplement.
%  \vspace*{-1em}
%\begin{acks}
%% MANUEL: Blind, so no acknowledgments yet.
%\paragraph*{Acknowledgements.} Blind
% M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
%\manuel{}{: We should also thank the COST ACTION: Ask Carlos?} \ekhine{}{Sure!, esto no habra q ponerlo aun por ser blind, no?}
%Thanks to: Hao Wang (Leiden University) for pointing us to the arguments of
%\citep{EriPeaGar2019scalable}; Thomas Stützle (Université libre de Bruxelles) for suggesting works
%about differences between QAPLIB instances; and Ahmed Boujaada for the experimental support. 
  % supported by the \grantsponsor{GS501100001809}{National Natural
  %   Science Foundation of
  %   China}{http://dx.doi.org/10.13039/501100001809} under Grant
  % No.:~\grantnum{GS501100001809}{61273304}
  % and~\grantnum[http://www.nnsf.cn/youngscientists]{GS501100001809}{Young
  %   Scientists' Support Program}.
%\end{acks}
